Dynamical modeling of tidal streams

### 2014-01-02 tex/ms.tex
streams: tweaks after reading; first draft

However, streams are observed in position--velocity coordinates and
for action--angle descriptions to be useful, we must be able to
calculate the transformation between these coordinate systems
efficiently. Until now this has required accurate phase--space data
and specialized algorithms to calculate action--angle coordinates that
break down for the eccentric orbits on which tidal-stream progenitors
are typically found (that is, radial and/or vertical actions of
similar size as the angular momentum; \citealt{Sanders12a}). However,
even with Gaia, future stream data will typically have large
uncertainties compared to the intrinsic dispersion of the stream and
in particular the line-of-sight velocities of the faintest stream
members will not be observed for many stars. Additionally, streams are
superimposed on a non-negligible background of field stars and
background contamination cannot be easily taken into account in any of
the current stream-fitting methods.

In this paper, I present a new method for modeling cold tidal streams
that fundamentally lives in frequency--angle space, because the stream
distribution function is essentially one-dimensional in this
space. While many of the results of this paper are more generally
valid, the fiducial stream model in this paper consists of a
three-dimensional (close to) Gaussian distribution of frequencies in
the stream, a three-dimensional Gaussian distribution of initial angle
offsets between stream members and the progenitor, and a uniform
distribution of stripping times; the frequency and angle distributions
are independent of stripping time. While this model oversimplifies the
real stripping process, it is argued that at least for constraining
the gravitational potential with stream data, this simple model is
likely unbiased. For detailed mock stream data, more elaborate models
of the stripping process as a function of time might be necessary and
I discuss how these could easily be incorporated into the proposed
framework.

To evaluate the model in position--velocity coordinates, the
transformation to and from action--angle coordinates is computed using
a novel method for calculating action--angle coordinates presented in
Appendix~\ref{sec:aa}. This transformation is linearized at a small
number of points along the progenitor orbit that span the width of the
stream. For cold streams, the separation between the stream track and
the progenitor orbit as well as the internal differences within the
stream are small, such that this linear approximation is accurate
enough. The linear approximation allows for fast evaluation of the
average stream location (the stream ``track'') in position--velocity
coordinates and efficient mock data generation. For a uniform
distribution of stripping times, an observed stream member's
phase--space probability distribution function (PDF) can be
analytically marginalized over stripping time, such that using the
linear approximation any evaluation---including marginalization and
uncertainty convolution---of the PDF is extremely rapid. This clears
the way for proper probabilistic inference of the Milky Way halo's
gravitational potential and progenitor properties using stream data
for individual stars.

@@

The dynamics of tidal streams both in real (position--velocity) and
action--angle space has been described by various previous authors
\citep[\eg,][]{Helmi99a,Tremaine99a,Johnston99a,Sanders13b}. I provide
here a brief description of the dynamics of tidal streams that is
relevant for what follows.

In real space, a tidal stream forms as stars are stripped from a
progenitor cluster or satellite galaxy some time $\ts$ in the
past and then evolve (largely) independently in a galaxy's host
potential, moving away both in position and velocity from the
progenitor object as time goes on. Thus, a star was offset by a small
amount $(\Delta \vecx,\Delta \vecv)$ from the position of the
progenitor $(\vecx^p,\vecv^p)(-\ts)$ at that time. Then the star
orbited under the influence of the same gravitational potential $\Phi$
as the progenitor until the present day $(t=0)$, when it was observed:
$(\vecx,\vecv)(t=0)$ = $\vech_{\ts}(\vecx^p(-ts)+\Delta
\vecx,\vecv^p(-\ts)+\Delta \vecv)$, where $\vech$ denotes the
Hamiltonian flow from $t=-\ts$ to $t=0$ (see
\citealt{binneytremaine}); this is simply the orbit of the star
between these times. In phase-space this orbit can be calculated by
solving Hamilton's equations: $\dot{\vecx} = \vecv; \dot{\vecv} = -
\dd \Phi / \dd \vecx$.

In action--angle space the dynamics of stream formation is much
simpler. A star received an offset ($\Delta \vecj,\Delta \veca$) at
time $-\ts$ from the action--angle coordinates of the progenitor
$(\vecj^p,\veca^p)$, similar to the small offset ($\Delta \vecx,\Delta
\vecv)$ in the real--space description. The offset in the actions
corresponds to an offset in the frequencies $\Delta \veco$;
ultimately, this frequency offset is responsible for the spreading of
the stream over long stretches on the sky. Because both the
progenitor's angles and the star's angles increase linearly with time,
albeit at different frequencies, the difference in the angles also
increases linearly in time. Therefore at time $t=0$ the difference in
angles is
\begin{equation}
\Delta \veca(t=0) = \Delta \veca(-\ts) + \Delta \veco \,\ts\,,
\end{equation}
while the difference in actions is constant
\begin{equation}
\Delta \vecj(t=0) = \Delta \vecj(-\ts)\,.
\end{equation}
As the initial offset $\Delta \veca(-\ts)$ in angles is small compared
to the growth $\Delta \veco \,\ts$, we have that $\Delta \veca(t=0)
\approx \Delta \veco \,\ts$.

Thus, in action--angle coordinates, the dynamics of tidal-stream
formation is simplified to linear growth of small initial angle
differences due to small action/frequency differences. The angle
direction $\unitvec{\veco} \equiv \Delta \veco / | \Delta \veco|$
along which a star moves away from the progenitor is related to the
initial action offset by $\Delta \veco \approx (\partial \veco /
\partial \vecj) \Delta \vecj = (\partial^2 H / \partial \vecj \partial
\vecj) \Delta \vecj$, in the limit of small action offsets, because
$\veco = \partial H / \partial \vecj$, where $H$ is the
Hamiltonian. For an approximately one-dimensional stream to form, the
Hessian matrix $(\partial^2 H / \partial \vecj \partial
\vecj)\large|_{\vecj^p}$ has to be dominated by a single large
eigenvalue; in this case the angle difference of stars stripped from
the stream at any time will fall along approximately the same
direction $\unitvec\veco$, and the stream will be essentially one
dimensional. In detail, because the distribution of initial offsets
$\Delta \vecj$ is not isotropic, the direction $\unitvec \veco$ is not
just determined by the potential (through $H$) and the progenitor
orbit, but also by the distribution of $\Delta \vecj$, which is set
through a combination of the internal dynamics of the progenitor and
the stripping process. For dynamical modeling we can be agnostic about
the relative importance of all of these contributors to $\unitvec
\veco$ by flexible modeling of the distribution of $\Delta \vecj$ or
$\Delta \veco$.

@@

I illustrate the theoretical considerations set forth in this paper by
using a simulated tidal stream. This stream is set up similarly to
that used in \citet{Sanders13b}. The stream is constructed by running
an $N$-body simulation of a King cluster on an orbit similar to that
of the GD-1 stream \citep{Koposov10a}. Specifically, a King cluster
with a mass of $2\times10^4\msun$, a tidal radius of $0.07\kpc$, and a
ratio of the central potential to the velocity dispersion squared of
$2.0$ is sampled using $10^4$ particles. The cluster is evolved in an
external logarithmic host potential with $V_c = 220\kms$ and a
flattening $q=0.9$ ($\Phi = V_c^2 \ln [R^2 + z^2/q^2]$) using the
gyrfalcon code \citep{Dehnen00a,Dehnen02a} with a softening of
$1.5\pc$ in the NEMO toolkit \citep{Teuben95a}. The cluster is evolved
for $5.011\Gyr$ ($=5.125\kpc/[\kms]$) from the initial condition
$(X,Y,Z) = (-11.63337239,-20.76235661,-10.631736273934635)\kpc$ and
$(V_X,V_Y,V_Z) = (-128.8281653,42.88727925,79.172383882274971)\kms$,
chosen such that the cluster ends up at roughly the observed location
of the GD-1 stream at the end of the simulation
\citep{Koposov10a}. The progenitor cluster's orbit has an eccentricity
of $0.31$, a pericenter of $13.7\kpc$, and a current $r=14.4\kpc$,
such that the cluster is close to pericenter at the end of the
simulation. The actions of the progenitor are $(J_R,L_Z,J_Z) =
(288.5,-3173.7,897.6)\kpc\kms$ and its orbital frequencies are
$(\Omega_R,\Omega_\phi,\Omega_Z) = (15.7,-10.8,11.9)\Gyr\inv$.

@@

A generative model of a tidal stream in frequency--angle space
requires three ingredients: (a) a model for the distribution of times
at which stars are stripped from the progenitor, (b) a prescription
for the distribution of frequency offsets from the progenitor at every
given stripping time, and (c) a description of the angle offsets for
any given frequency offset and stripping time. Having specified these
ingredients, we can then generate tidal debris at any given stripping
time and evolve it forward using the simple linear dynamics in
frequency--angle space discussed in
\sectionname~\ref{sec:dynamics}. The initial angle offsets are small
enough that they are likely unobservable even with futuristic data
(especially in the direction of the frequency offset, where the
initial angle offset is quickly overwhelmed by the subsequent
dynamical evolution; for the simulated stream used here this happens
after approximately $20\Myr$). For that reason, I will assume that the
initial angle offsets are independent of both stripping time and
frequency offset. In what follows, we will model them using a simple
isotropic Gaussian distribution.

\begin{figure}[t!]
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_timeshist.ps}
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_timesdO.ps}
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_timesda.ps}
  \caption{Properties of the simulated tidal stream of
    \sectionname~\ref{sec:sim} as a function of time. Stripping times
    are calculated by linear regression of the angle difference
    between a stream particle and the progenitor versus the equivalent
    frequency difference. The distribution of stripping times is shown
    in the left panel.  While the stripping process happens in bursts
    with a period of roughly $400\Myr$ (the radial period of the
    progenitor), it can be approximated as constant (more
    coarsely-binned solid histogram). The magnitude of the frequency
    offset between stream particles and the progenitor is shown in the
    middle panel. While particles stripped at pericenter have larger
    frequency offsets than those between pericenter passages, there is
    no secular trend and the typical frequency offset (averaged over a
    radial period) can be considered constant with time. The right
    panel shows the magnitude of the angle offset between stream
    particles and the progenitor as a function of stripping
    time. There is a spread in the angle offset at all stripping times
    because of the dispersion in frequency offsets, but typically
    stars that have been stripped earlier reach larger angle
    offsets. That is, self-sorting happens only to a small
    degree.}\label{fig:gd1_times}
%python plot_stream.py ../tex/gd1_evol_timeshist.ps
%python plot_stream.py ../tex/gd1_evol_timesdO.ps
%python plot_stream.py ../tex/gd1_evol_timesda.ps
\end{figure}

@@

The right panel of \figurename~\ref{fig:gd1_times} shows the angle
offset (at the end of the simulation) versus the stripping time. There
is a spread in the angle offset reach for any given stripping time,
but it is nevertheless typically the case that particles that have
been stripped earlier are found at larger angle offsets
today. Self-sorting---the tendency of particles stripped later with
larger frequency offsets to overtake particles stripped earlier at
smaller frequency differences and thus to erase correlations between
particle position along the stream and stripping time---happens to a
small extent but the narrow distribution of frequency offsets limits
the ability of particles removed later to reach large distances from
the progenitor.

We can model streams in frequency space because the mapping $\vecj
\rightarrow \veco$ is of full rank for axisymmetric and triaxial
potentials. However, for spherical potentials this mapping only has
rank two, because $\Omega_\phi = \mathrm{sgn}(L_Z)\,\Omega_Z$, such
that the determinant $|\partial \veco / \partial \vecj|$ is zero. For
spherical potentials, tidal streams only spread in a five-dimensional
subspace and in modeling stream in spherical potentials we need to
keep the direction perpendicular to this subspace constant.

@@

First, we approximate the bursty, non-uniform distribution of
stripping times in \figurename~\ref{fig:gd1_times} with a uniform
distribution
\begin{align}\label{eq:pt}
  \begin{split}
    p(\ts) & \propto 1\,,\qquad 0 < \ts < t_d\\
    & = 0\,,\qquad \mathrm{otherwise}\,.
  \end{split}
\end{align}
The thick, solid histogram in the right panel of
\figurename~\ref{fig:gd1_times} shows a coarser-binned histogram of
stripping times, indicating that the uniform distribution is a good
approximation over intervals larger than a radial period ($\approx
400\Myr$). The \emph{disruption time} $t_d$ is a free parameter of the
model that needs to be determined for each stream. For the model of
the simulated stream used here I set $t_d = 4.5\Gyr$.

@@

\begin{figure}[t!]
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_aadohist.ps}
  \caption{Distribution of the magnitude of frequency offsets between
    stream particles and the progenitor along the direction of the
    mean frequency offset. The dashed line is a Gaussian fit to this
    distribution, while the solid line is a fit of our preferred form
    for this distribution, given in \equationname~(\ref{eq:do}). Both
    forms provide adequate fits to the
    distribution. }\label{fig:gd1_dohist}
%python plot_stream.py ../tex/gd1_evol_aadohist.ps
\end{figure}

I then further constrain the mean frequency offset to lie along the
principal eigenvector of $V_\veco$. This mean frequency offset can be
described in terms of a parameter $\mu_\Omega = \Delta \Omega^m /
\sigma_{\Omega,1}$, where $\sigma_{\Omega,1}$ is the square root of
the largest eigenvalue of $V_\veco$ and $\Delta \Omega^m$ is the
one-dimensional mean frequency offset. The sign of this parameter sets
whether we are modeling a leading or a trailing
arm. \figurename~\ref{fig:gd1_dohist} shows the distribution of
frequency offsets in the direction of the median frequency offset in
the stream. The dashed line is a Gaussian fit to this distribution,
while the solid line is a fit of the form
\begin{equation}\label{eq:do}
  p(|\Delta \veco|_\parallel) = |\Delta \veco|_\parallel\,\mathcal{N}\left(|\Delta \veco|_\parallel|\Delta \Omega^m,\sigma_{\Omega,1}^2\right)\,,
\end{equation}
where $\mathcal{N}(\cdot|\cdot,\cdot)$ represents a Gaussian
distribution. The fit is equally good (because the dispersion is much
smaller than the mean) and I choose the second form because it
simplifies some of the stream-track calculations below. The best-fit
$(\Delta \Omega^m,\sigma_{\Omega,1})$ is $(0.19,0.033)\Gyr\inv$ in
both cases. Based on this information, a model with $\sigv =
0.365\kms$ and $\mu_\Omega = 6$ provides a good fit to the simulation
data. The distribution of frequency offsets in the two-dimensional
space perpendicular to the principal eigenvector of $V_\veco$ is
modeled as a zero-mean Gaussian with variance matrix given by the
projection of $V_\veco$ onto this space.

@@

The fiducial model for a tidal stream in frequency--angle coordinates
proposed here is therefore described by essentially two free
parameters in addition to the progenitor's phase-space position: the
progenitor's velocity dispersion \sigv\ and the disruption time
$t_d$. As shown below, these two parameters are the most important in
determining the track of the stream. More sophisticated models can
leave $\mu_\Omega$ and $\sigma_{\theta}$ as free parameters as
well. In what follows, I will fix these to the values given in the
previous paragraph. It must be stressed, however, that all of these
parameters here have been fixed ``by-eye'', and better fits might be
possible using more quantitative fit procedures.

@@

In this section, I discuss how to calculate the track of a model tidal
stream in the generative model described in
\sectionname~\ref{sec:generative}. The track consists of the mean
location of the stream as a function of angle along the stream. I also
describe how to estimate the dispersion of the stream along this
track. In \sectionname~\ref{sec:trackaa}, I explain how to calculate
the track of the stream in frequency--angle coordinates as well as how
to estimate the spread around this track. In
\sectionname~\ref{sec:trackxv} I discuss how to efficiently project
the stream track and dispersion into position--velocity coordinates
$(\vecx,\vecv)$ or observable coordinates $(l,b,D,\vlos,\pmll,\pmbb)$.

@@

\begin{figure}[t!]
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_aaaparopar.ps}
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_aaaparoperp.ps}\\
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_aaaparaperp.ps}
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_aaapartime.ps}
  \caption{Stream properties as a function of angle along the
    stream. The top, left panel shows the magnitude of frequency
    offsets along the stream. The gray line shows the mean predicted
    frequency offset calculated using the framework of
    \sectionname\sectionname~\ref{sec:modeloa} and \ref{sec:track},
    while the grayscale bands show the $68\,\%$ and $95\,\%$ limits of
    the predicted distribution. The top, right panel shows the
    magnitude of frequency offsets perpendicular to those in the left
    panel. The predicted distribution is constant as a function of
    angle and therefore not shown. The bottom, left panel shows the
    angle offset perpendicular to the stream and the grayscale bands
    show $68\,\%$ and $95\,\%$ limits of the predicted distribution of
    angle offsets; the white dashed line shows the $68\,\%$ limit
    calculated using a simplified method. The bottom, right panel
    shows the stripping time. The gray line and grayscale bands again
    show the mean and $68\,\%$ and $95\,\%$ confidence limits of the
    predicted distribution. While the ``burstiness'' related to the
    energetic stripping at pericenter that is apparent in the
    simulated data's distributions is (by design) absent in the model,
    the overall scale and the mean trends with angle are well
    represented by the model.}\label{fig:gd1_apar}
%python plot_stream.py ../tex/gd1_evol_aaaparopar.ps
%python plot_stream.py ../tex/gd1_evol_aaaparoperp.ps
%python plot_stream.py ../tex/gd1_evol_aaaparaperp.ps
%python plot_stream.py ../tex/gd1_evol_aaapartime.ps
\end{figure}

>>>>

However, streams are observed in position--velocity coordinates and
for action--angle descriptions to be useful, we must be able to
calculate the transformation between these coordinate systems
efficiently. Until now this has required accurate phase--space data
and specialized algorithms that break down for the eccentric orbits on
which tidal-stream progenitors are typically found (that is, radial
and/or vertical actions of similar magnitude as the angular momentum;
\citealt{Sanders12a}). However, even with Gaia, future stream data
will typically have large uncertainties compared to the intrinsic
dispersion of the stream and in particular the line-of-sight
velocities of the faintest stream members will not be observed for
many stars. Additionally, streams are superimposed on a non-negligible
background of field stars and background contamination cannot be
easily taken into account in any of the current stream-fitting
methods.

In this paper, I present a new method for modeling cold tidal streams
that fundamentally lives in frequency--angle space, because the stream
distribution function is essentially one-dimensional in this
space. While many of the results of this paper are more generally
valid, the fiducial stream model in this paper consists of a
three-dimensional (close to) Gaussian distribution of frequencies in
the stream, a three-dimensional Gaussian distribution of initial angle
offsets between stream members and the progenitor, and a uniform
distribution of stripping times; the frequency and angle distributions
are independent of stripping time. While this model oversimplifies the
real stripping process, I argue that at least for constraining the
gravitational potential with stream data, this simple model is likely
unbiased. For detailed mock stream data, more elaborate models of the
stripping process as a function of time might be necessary and I
discuss how these could easily be incorporated into the proposed
framework.

To evaluate the model in position--velocity coordinates, the
transformation to and from action--angle coordinates is computed using
a novel method for calculating action--angle coordinates presented in
Appendix~\ref{sec:aa}. This transformation is linearized at a small
number of points along the progenitor orbit that span the length of
the stream. For cold streams, the separation between the stream track
and the progenitor orbit as well as the internal differences within
the stream are small, such that this linear approximation is accurate
enough. The linear approximation allows for fast evaluation of the
average stream location (the stream ``track'') in position--velocity
coordinates and efficient mock data generation. For a uniform
distribution of stripping times, an observed stream member's
phase--space probability distribution function (PDF) can be
analytically marginalized over stripping time, such that using the
linear approximation any evaluation---including marginalization and
uncertainty convolution---of the PDF is extremely rapid. This clears
the way for proper probabilistic inference of the Milky Way halo's
gravitational potential and progenitor properties using stream data
for individual stars.

@@

The dynamics of tidal streams both in position--velocity and
action--angle space has been described previously by various authors
\citep[\eg,][]{Helmi99a,Tremaine99a,Johnston99a,Sanders13b}. I provide
here a brief description of the dynamics of tidal streams that is
relevant for what follows.

In position--velocity space, a tidal stream forms as stars are
stripped from a progenitor cluster or satellite galaxy some time $\ts$
in the past and then evolve (largely) independently in a galaxy's host
potential, moving away both in position and velocity from the
progenitor object as time goes on. Thus, a star was offset by a small
amount $(\Delta \vecx,\Delta \vecv)$ from the position of the
progenitor $(\vecx^p,\vecv^p)(-\ts)$ at that time. Then the star
orbited under the influence of the same gravitational potential $\Phi$
as the progenitor until the present day $(t=0)$, when it was observed:
$(\vecx,\vecv)(t=0)$ = $\vech_{\ts}(\vecx^p(-ts)+\Delta
\vecx,\vecv^p(-\ts)+\Delta \vecv)$, where $\vech$ denotes the
Hamiltonian flow from $t=-\ts$ to $t=0$ (see
\citealt{binneytremaine}); this is simply the orbit of the star
between these times. In phase-space this orbit can be calculated by
solving Hamilton's equations: $\dot{\vecx} = \vecv; \dot{\vecv} = -
\dd \Phi / \dd \vecx$.

In action--angle space the dynamics of stream formation is much
simpler. A star received an offset ($\Delta \vecj,\Delta \veca$) at
time $-\ts$ from the action--angle coordinates of the progenitor
$(\vecj^p,\veca^p)$, similar to the small offset ($\Delta \vecx,\Delta
\vecv)$. The offset in the actions corresponds to an offset in the
frequencies $\Delta \veco$; ultimately, this frequency offset is
responsible for the spreading of the stream over long stretches on the
sky. Because both the progenitor's angles and the star's angles
increase linearly with time, albeit at different frequencies, the
difference in the angles also increases linearly in time. Therefore at
time $t=0$ the difference in angles is
\begin{equation}
\Delta \veca(t=0) = \Delta \veca(-\ts) + \Delta \veco \,\ts\,,
\end{equation}
while the difference in actions is constant
\begin{equation}
\Delta \vecj(t=0) = \Delta \vecj(-\ts)\,.
\end{equation}
As the initial offset $\Delta \veca(-\ts)$ in angles is small compared
to the growth $\Delta \veco \,\ts$, we have that $\Delta \veca(t=0)
\approx \Delta \veco \,\ts$.

Thus, in action--angle coordinates, the dynamics of tidal-stream
formation is simplified to linear growth of small initial angle
differences due to small action/frequency differences. The angle
direction $\unitvec{\veco} \equiv \Delta \veco / | \Delta \veco|$
along which a star moves away from the progenitor is related to the
initial action offset by $\Delta \veco \approx (\partial \veco /
\partial \vecj) \Delta \vecj = (\partial^2 H / \partial \vecj \partial
\vecj) \Delta \vecj$, in the limit of small action offsets, because
$\veco = \partial H / \partial \vecj$, where $H$ is the
Hamiltonian. For an approximately one-dimensional stream to form, the
Hessian matrix $(\partial^2 H / \partial \vecj \partial
\vecj)\large|_{\vecj^p}$ has to be dominated by a single large
eigenvalue; in this case the angle difference of stars stripped from
the stream at any time will fall along approximately the same
direction $\unitvec\veco$, and the stream will be essentially one
dimensional. In detail, because the distribution of initial offsets
$\Delta \vecj$ is not isotropic, the direction $\unitvec \veco$ is not
just determined by the potential (through $H$) and the progenitor
orbit $\vecj^p$, but also by the distribution of $\Delta \vecj$, which
is set through a combination of the internal dynamics of the
progenitor and the stripping process. For dynamical modeling we can be
agnostic about the relative importance of all of these contributors to
$\unitvec \veco$ by flexible modeling of the distribution of $\Delta
\vecj$ or $\Delta \veco$.

@@

I illustrate the theoretical considerations set forth in this paper by
using a simulated tidal stream. This stream is set up similarly to
that used in \citet{Sanders13b}. The stream is constructed by running
an $N$-body simulation of a King cluster on an orbit similar to that
of the GD-1 stream \citep{Koposov10a}. Specifically, a King cluster
with a mass of $2\times10^4\msun$, a tidal radius of $0.07\kpc$, and a
ratio of the central potential to the velocity dispersion squared of
$2.0$ is sampled using $10^4$ particles. The cluster is evolved with
self-gravity in an external logarithmic host potential with $V_c =
220\kms$ and a flattening $q=0.9$ ($\Phi = V_c^2 \ln [R^2 + z^2/q^2]$)
using the gyrfalcon code \citep{Dehnen00a,Dehnen02a} with a softening
of $1.5\pc$ in the NEMO toolkit \citep{Teuben95a}. The cluster is
evolved for $5.011\Gyr$ ($=5.125\kpc/[\kms]$) from the initial
condition $(X,Y,Z) =
(-11.63337239,-20.76235661,-10.631736273934635)\kpc$ and
$(V_X,V_Y,V_Z) = (-128.8281653,42.88727925,79.172383882274971)\kms$,
chosen such that the cluster ends up at roughly the observed location
of the GD-1 stream at the end of the simulation
\citep{Koposov10a}. The progenitor cluster's orbit has an eccentricity
of $0.31$, a pericenter of $13.7\kpc$, and a current $r=14.4\kpc$,
such that the cluster is close to pericenter at the end of the
simulation. The actions of the progenitor are $(J_R,L_Z,J_Z) =
(288.5,-3173.7,897.6)\kpc\kms$ and its orbital frequencies are
$(\Omega_R,\Omega_\phi,\Omega_Z) = (15.7,-10.8,11.9)\Gyr\inv$.

@@

A generative model of a tidal stream in frequency--angle space
requires three ingredients: (a) a model for the distribution of times
at which stars are stripped from the progenitor, (b) a prescription
for the distribution of frequency offsets from the progenitor at every
given stripping time, and (c) a description of the angle offsets for
any given frequency offset and stripping time. Having specified these
ingredients, we can then generate tidal debris at any given stripping
time and evolve it forward using the simple linear dynamics in
frequency--angle space discussed in
\sectionname~\ref{sec:dynamics}. The initial angle offsets are small
enough that they are likely unobservable even with futuristic data
(especially in the direction of the frequency offset, where the
initial angle offset is quickly overwhelmed by the subsequent
dynamical evolution; for the simulated stream used here this happens
after approximately $20\Myr$). For that reason, I will assume that the
initial angle offsets are independent of both stripping time and
frequency offset. In what follows, I will model them using a simple
isotropic Gaussian distribution.

\begin{figure}[t!]
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_timeshist.ps}
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_timesdO.ps}
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_timesda.ps}
  \caption{Properties of the simulated tidal stream of
    \sectionname~\ref{sec:sim} as a function of time. Stripping times
    are calculated by linear regression of the angle difference
    between a stream particle and the progenitor versus the equivalent
    frequency difference. The distribution of stripping times is shown
    in the left panel.  While the stripping process happens in bursts
    with a period of roughly $400\Myr$ (the radial period of the
    progenitor), it can be approximated as constant (more
    coarsely-binned solid histogram). The magnitude of the frequency
    offset between stream particles and the progenitor is shown in the
    middle panel. While particles stripped at pericenter have larger
    frequency offsets than those between pericenter passages, there is
    almost no secular trend and the typical frequency offset (averaged
    over a radial period) can be considered constant with time. The
    right panel shows the magnitude of the angle offset between stream
    particles and the progenitor as a function of stripping
    time. There is a spread in the angle offset at all stripping times
    because of the dispersion in frequency offsets, but typically
    stars that have been stripped earlier reach larger angle
    offsets. That is, self-sorting happens only to a small
    degree.}\label{fig:gd1_times}
%python plot_stream.py ../tex/gd1_evol_timeshist.ps
%python plot_stream.py ../tex/gd1_evol_timesdO.ps
%python plot_stream.py ../tex/gd1_evol_timesda.ps
\end{figure}

@@

The right panel of \figurename~\ref{fig:gd1_times} shows the angle
offset (at the end of the simulation) versus the stripping time. There
is a spread in the angle offset reached for any given stripping time,
but it is nevertheless typically the case that particles that have
been stripped earlier are found at larger angle offsets
today. Self-sorting---the tendency of particles stripped later with
larger frequency offsets to overtake particles stripped earlier at
smaller frequency differences and thus to erase correlations between
particle position along the stream and stripping time---happens to a
small extent but the narrow distribution of frequency offsets limits
the ability of particles removed later to reach large distances from
the progenitor.

We can model streams in frequency space because the mapping $\vecj
\rightarrow \veco$ is of full rank for axisymmetric and triaxial
potentials. However, for spherical potentials this mapping only has
rank two, because $\Omega_\phi = \mathrm{sgn}(L_Z)\,\Omega_Z$, such
that the determinant $|\partial \veco / \partial \vecj|$ is zero. For
spherical potentials, tidal streams only spread in a five-dimensional
subspace and in modeling streams in spherical potentials we need to
keep the direction perpendicular to this subspace constant.

@@

First, we approximate the bursty, non-uniform distribution of
stripping times in \figurename~\ref{fig:gd1_times} with a uniform
distribution
\begin{align}\label{eq:pt}
  \begin{split}
    p(\ts) & \propto 1\,,\qquad 0 < \ts < t_d\\
    & = 0\,,\qquad \mathrm{otherwise}\,.
  \end{split}
\end{align}
The thick, solid histogram in the right panel of
\figurename~\ref{fig:gd1_times} shows a coarser-binned histogram of
stripping times, indicating that the uniform distribution is a good
approximation over intervals longer than a radial period ($\approx
400\Myr$). The \emph{disruption time} $t_d$ is a free parameter of the
model that needs to be determined for each stream. For the model of
the simulated stream used here I set $t_d = 4.5\Gyr$.

@@

\begin{figure}[t!]
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_aadohist.ps}
  \caption{Distribution of the magnitude of frequency offsets between
    stream particles and the progenitor along the direction of the
    mean frequency offset. The dashed line is a Gaussian fit to this
    distribution, while the solid line is a fit of our preferred form
    for this distribution, given in \equationname~(\ref{eq:do}) (the
    dashed line is almost entirely covered by the solid line). Both
    forms provide adequate fits to the
    distribution. }\label{fig:gd1_dohist}
%python plot_stream.py ../tex/gd1_evol_aadohist.ps
\end{figure}

I then further constrain the mean frequency offset to lie along the
principal eigenvector of $V_\veco$. This mean frequency offset can be
described in terms of a parameter $\mu_\Omega = \Delta \Omega^m /
\sigma_{\Omega,1}$, where $\sigma_{\Omega,1}$ is the square root of
the largest eigenvalue of $V_\veco$ and $\Delta \Omega^m$ is the
one-dimensional mean frequency offset. The sign of $\mu_\Omega$ sets
whether we are modeling a leading or a trailing
arm. \figurename~\ref{fig:gd1_dohist} shows the distribution of
frequency offsets in the direction of the median frequency offset in
the stream. The dashed line is a Gaussian fit to this distribution,
while the solid line is a fit of the form
\begin{equation}\label{eq:do}
  p(|\Delta \veco|_\parallel) = |\Delta \veco|_\parallel\,\mathcal{N}\left(|\Delta \veco|_\parallel|\Delta \Omega^m,\sigma_{\Omega,1}^2\right)\,,
\end{equation}
where $\mathcal{N}(\cdot|\cdot,\cdot)$ represents a Gaussian
distribution. The fit is equally good (because the dispersion is much
smaller than the mean) and I choose the second form because it
simplifies some of the stream-track calculations below. The best-fit
$(\Delta \Omega^m,\sigma_{\Omega,1})$ is $(0.19,0.033)\Gyr\inv$ in
both cases. Based on this information, a model with $\sigv =
0.365\kms$ and $\mu_\Omega = 6$ provides a good fit to the simulation
data. The distribution of frequency offsets in the two-dimensional
space perpendicular to the principal eigenvector of $V_\veco$ is
modeled as a zero-mean Gaussian with variance matrix given by the
projection of $V_\veco$ onto this space.

@@

The fiducial model for a tidal stream in frequency--angle coordinates
proposed here is therefore described by essentially two free
parameters in addition to the progenitor's phase-space position: the
progenitor's velocity dispersion \sigv\ and the disruption time
$t_d$. As shown below, these two parameters are the most important in
determining the track of the stream. More sophisticated models can
leave $\mu_\Omega$ and $\sigma_{\theta}$ as free parameters as
well. In what follows, I will fix these to the values given in the
previous paragraphs. It must be stressed, however, that all of these
parameters here have been fixed ``by-eye'', and better fits might be
possible using more quantitative fit procedures.

@@

In this Section, I discuss how to calculate the track of a model tidal
stream in the generative model described in
\sectionname~\ref{sec:generative}. The track consists of the mean
location of the stream as a function of angle along the stream. I also
describe how to estimate the dispersion of the stream along this
track. In \sectionname~\ref{sec:trackaa}, I explain how to calculate
the track of the stream in frequency--angle coordinates as well as how
to estimate the spread around this track. In
\sectionname~\ref{sec:trackxv} I discuss how to efficiently project
the stream track and dispersion into position--velocity coordinates
$(\vecx,\vecv)$ or observable coordinates $(l,b,D,\vlos,\pmll,\pmbb)$.

@@

\begin{figure}[t!]
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_aaaparopar.ps}
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_aaaparoperp.ps}\\
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_aaaparaperp.ps}
  \includegraphics[width=0.32\textwidth,clip=]{gd1_evol_aaapartime.ps}
  \caption{Stream properties as a function of angle along the
    stream. The top, left panel shows the magnitude of frequency
    offsets along the stream. The gray line shows the mean predicted
    frequency offset calculated using the framework of
    \sectionname\sectionname~\ref{sec:modeloa} and \ref{sec:track},
    while the grayscale bands show the $68\,\%$ and $95\,\%$ limits of
    the predicted distribution. The dashed line shows the angle $\apar
    = \Delta \Omega^m\,t_d$, angles beyond which can only be reached
    by having a larger $\Delta \opar$ than average. The top, right
    panel shows the magnitude of frequency offsets perpendicular to
    those in the left panel. The predicted distribution is constant as
    a function of angle and therefore not shown. The bottom, left
    panel shows the angle offset perpendicular to the stream and the
    grayscale bands show $68\,\%$ and $95\,\%$ limits of the predicted
    distribution of angle offsets; the white dashed line shows the
    $68\,\%$ limit calculated using a simplified method. The bottom,
    right panel shows the stripping time. The gray line and grayscale
    bands again show the mean and $68\,\%$ and $95\,\%$ confidence
    limits of the predicted distribution. While the ``burstiness''
    related to the energetic stripping at pericenter that is apparent
    in the simulated data's distributions is (by design) absent in the
    model, the overall scale and the mean trends with angle are well
    represented by the model.}\label{fig:gd1_apar}
%python plot_stream.py ../tex/gd1_evol_aaaparopar.ps
%python plot_stream.py ../tex/gd1_evol_aaaparoperp.ps
%python plot_stream.py ../tex/gd1_evol_aaaparaperp.ps
%python plot_stream.py ../tex/gd1_evol_aaapartime.ps
\end{figure}
